Precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-26 manuscript.tex
introduction

Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} present in the neural code is essential in understanding information processing and yet, most neural models do not take advantage of this fine-scaled temporal dimension. This hypothesis is directly inspired by neurobiological observations, particularly in the hippocampus, and it expands the capabilities of analog representations (which are based on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. 
Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and also suggest the existence of repeated spatio-temporal spiking motifs in the spike train reaching neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Importantly, validating this hypothesis would also be crucial in our understanding of neuroscientific processes.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a review of the state of the art on the role of dynamics in vision. Vision is the set of processes that allow us to make sense of the luminous world, and is an area intensively studied in neuroscience with respect to the question of neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between external visual space and its representation on the retina. The origins of this  neuroscientific question can be found in the first experiments of Pierre Flouren showing the relationship between sensations, perception and activity in the cerebral cortex using lesions in animals~\citep{pearce_marie-jean-pierre_2009}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as orientation or motion. Remarkably, there is a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

In particular, there have been some remarkable findings when studying the dynamics of vision. Indeed, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. Its first experiment consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges at a very early latency in neural activity. These results have been extended to several species including primates but also extended to different experimental protocols and have shown for example that the response could be extremely fast of the order of $120~\ms$ when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}. This fast processing also explains the surprising experiments of fast serial detection, which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal~\citep{keysers_speed_2001}. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only $14~\ms$ per image.

This speed of the visual cortex in primates, although surprising, is  compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented to the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex, and it takes about $55~\ms$ in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.
%
An important consequence of the speed of processing of vision is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only $120~\ms$ consists of about ten processing stages following the ``forward'' pathways of the visual system, then this imposes that the processing in a single area is performed with a reduced number of spikes. Altogether, these results show that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.

>>>>

Indeed, the response of one biological neuron largely depends on the precise timing of the sequence of incoming spikes that reach its soma. In comparison to a classical analogical vector of inputs, this \emph{event-based representation} present in the neural code is essential in understanding information processing and yet, most neural models do not take advantage of this fine-scaled temporal dimension. This hypothesis is directly inspired by neurobiological observations, particularly in the hippocampus, and it expands the capabilities of analog representations (which are based on the firing rate) by considering a representation based on repetitions of spiking motifs at precise times of occurrence. 
Additionally, numerous studies demonstrated the importance of precise timing in neural population activity~\citep{davis_spontaneous_2021}, efficient encoding thanks to the use of spike latencies~\citep{perrinet_coding_2004,gollisch_rapid_2008} or precise timing in the auditory system~\citep{deweese_binary_2003,carr_circuit_1990}. All these findings, and more~\citep{bohte_evidence_2004}, highlight the importance of the temporal aspect of the neural code and also suggest the existence of repeated spatio-temporal spiking motifs in the spike train reaching neurons.  A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data.  Importantly, validating this hypothesis would also be crucial in our understanding of neuroscientific processes.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Precise timing in the neural code enables efficient vision}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
Let us start with a review of the state of the art on the role of dynamics in vision. Vision is the set of processes that allow us to make sense of the luminous world, and is an area intensively studied in neuroscience, in particular with respect to deciphering the neural code. In most mammals, light enters the eye to induce neural activity on the retina, which maintains a certain similitude between external visual space and its representation on the retina. The origins of this  neuroscientific question can be found in the first experiments of Pierre Flouren showing the relationship between sensations, perception and activity in the cerebral cortex using lesions in animals~\citep{pearce_marie-jean-pierre_2009}. In a series of seminal studies,~\citet{hubel_receptive_1968} showed that this activity could be selective to different features, such as orientation or motion. Remarkably, there is a monotonic relationship between the contrast of visual features and the firing frequency of neurons. However, there is no consensus to explain the multiple nonlinear mechanisms that transform the visual scene into retinotopic neural activity maps, even though these processes seem to constitute essential pieces to this puzzle~\citep{carandini_normalization_2012}.

In particular, there have been some remarkable findings when studying the dynamics of vision. For instance, Simon Thorpe's group has shown during the last decades numerous examples demonstrating that humans can categorize briefly presented images in a fraction of a second. Its first experiment consisted in asking subjects to categorize images that do or do not contain animals~\citep{thorpe_speed_1996}. The results showed that humans were able to perform this task very well (with a success rate of more than 95\%) but above all that a differential activity for the two categories of images could be observed by electroencephalography, showing that this differentiation emerges with a very short latency in neural activity. These results have been extended to several species including primates, and was also extended to different experimental protocols and have shown for example that the response could be extremely fast of the order of $120~\ms$ when the task was to perform a saccade~\citep{kirchner_ultra-rapid_2006}. This fast processing also explains the surprising experiments of fast serial detection, which consists in presenting a fast succession of different images and to decode via the EEG if the observer can detect for example the presence of an animal~\citep{keysers_speed_2001}. The performances decrease progressively as the frequency of presentation of the images increases. However, it has been shown in the macaque that a significant performance could be maintained with an image presentation time of only $14~\ms$ per image.

This speed of the visual cortex in primates, although surprising, is  compatible with the latencies that are recorded at the neuro-physiological level. Indeed, when an image is presented to the retina, the visual information is rapidly propagated to the thalamus and then to the primary visual cortex, and it takes about $55~\ms$ in the macaque~\citep{nowak_timing_1997}, see Figure~\ref{fig:thorpe}. This functioning of visual processing as a forward pass is most prominent in fast processing but can be complemented with feedback loops from the higher areas to the sensory areas~\citep{lamme_distinct_2000}.
%
An important consequence of the speed of processing of vision is that it implies that it is carried out using only very few spikes. Indeed, if we consider that a behavioral response in only $120~\ms$ consists of about ten processing stages following the ``forward'' pathways of the visual system. This was was indeed modeled in feed-forward models trained with back-propagation~\cite{serre_feedforward_2007,jeremie_ultrafast_2022}. This imposes that the processing in a single area is performed with a reduced number of spikes. Altogether, these results show that ``like other senses, vision relies heavily on temporal strategies and temporal neural codes to extract and represent spatial information''~\citep{rucci_temporal_2018}.
